Classless Java: Tuning Java Interfaces

### 2015-12-02 paper/sections/Overview.tex
revising text

\begin{figure}
\begin{tabular}{l|l}
\includegraphics[height=4cm]{PegasusDetail.pdf}\hspace{20pt} &
\begin{minipage}{8cm}
\vspace{-90pt}
\lstinputlisting[linerange=6-17]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_JAVA
\end{minipage}
\end{tabular}
\caption{The animal system. Complete animal system on the left, and code for simplified animal system on the right. }\label{fig:pegasus}
\end{figure}

@@

\vspace{-5pt}
\paragraph{\Q@Point2D@ with Object Interfaces}
This obvious constructor code can be automatically generated by our
\mixin annotation.  By annotating the interface \Q@Point2D@, a variation of the shown static method
\texttt{of} will be generated, mimicking the functionality of a simple-minded
constructor: by looking at the methods that need implementation it
first detects what are the fields, then generates an \Q@of@ method with
one parameter for each of them. That is, we can just write

@@

\noindent More precisely, a field or factory parameter is generated for every
no-args method requiring implementation (except for methods whose name have
special meaning).  An example of code using \Q@Point2D@ is:

\begin{lstlisting}
Point2D p = Point2D.of(42,myPoint.y());
\end{lstlisting}

\noindent where we return a new point, using \Q@42@ as x-coordinate,
and taking all the other informations (only \Q@y@ in this case) from
another point.

\paragraph{\texttt{with} methods in Object Interfaces}
The pattern of creating a new piece of data by reusing most
information from an old piece of data is very common
when programming with immutable data-structures; it is so common that
it is supported in our code generation as \Q@with@ methods,
that is:
\begin{lstlisting}
@Obj interface Point2D { 
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); // with methods }
\end{lstlisting}

\noindent Using \texttt{with} methods, the point \texttt{p} could have
been created instead using:

@@

\noindent If there is a large number of fields, \texttt{with} methods
will save programmers from writting large amounts of tedious code that
simply copies the values of fields. 
\begin{comment}
is expanded by our annotation into\footnote{
Note how we actually generate a real field \Q@int x=_x;@.
This provide a more uniform translation that can work also for mutable datastructures, where setters are required.
}

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
\end{comment}
Moreover, if the programmer wants a different implementation, they
just provide it using a \Q@default@ method. For example:
\begin{lstlisting}
@Obj interface Point2D{ 
  int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}

\noindent can be used by programmers to define their own behaviour for 
the \texttt{with} methods.

@@

\paragraph{\Q@Animal@ and \Q@Horse@: simple mutable data with fields}
Two dimensional points are mathematical entities, thus we chosen to
use immutable data structure to model them. However animals are real
world entities, and when an animal moves, it is the same animal that
now has a different location. We model this with mutable state.

%Now we proceed to define \texttt{Animal} with \texttt{point} ``member
%field''.
%%Not again, before we used only getters!
% Again, we model this member field with getter and setter methods:
\lstinputlisting[linerange=59-59]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=ANIMAL

\noindent Here we declare abstract getter and setter for the mutable ``field'' \Q@location@.
The \mixin annotation is not used. This is morally
equivalent to an abstract class in full Java; and there is no
convenient way to instantiate it.
For \texttt{Horse}, the \mixin annotation is used and a concrete implementation of \texttt{run()} method
is defined using a default method, which also further illustrates the convenience of \emph{with} methods.

\lstinputlisting[linerange=63-64]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=HORSE

\noindent Creating and using a horse animal is quite simple:

@@

\paragraph{Summary} Dealing with state (whether mutable or immutable)
in object interfaces relies on a notion of abstract state, where only
methods to interact with state are available to users. Object
interfaces provides support for four different types abstract state
operations, which are summarized in Figure~\ref{fig:abstractstate}.
The abstract state operations are determined by naming conventions and the
types of the methods. Fluent setters are a variant of mutable setters,
and are discussed in the case study in Section~\ref{sec:dsls}.

@@

\subsection{Object Interfaces and Subtyping}
\Q@Bird@s are \Q@Animal@s, but while \Q@Animal@s only need 2D
locations, \Q@Bird@s need 3D locations. Therefore when the \texttt{Bird}
interface extends the animal interface, the notion of points needs to
be refined. Such kind of refinement usually poses a challenge
in typical class-based approaches. Fortunatelly, with object interfaces,
we are able to provide a simple and effective solution to that problem.

\paragraph{Unsatisfactory class-based solutions to field type refinement}
In Java if we define an animal class with a field we have a set of unsatisfactory options in front of us:
\begin{itemize}
\item Define a \Q@Point3D@ field in \Q@Animal@:
this is bad since all animals would require more that is needed, and also it may requires the programmer to predict the future, or it may require to adapt the old code to accommodate for new evolutions.
\item Define a \Q@Point2D@ field in \Q@Animal@ and define an extra \Q@int z@ field in \Q@Bird@.
This solution is very ad-hoc, requires to basically duplicate the difference between \Q@Point2D@ and \Q@Point3D@ inside of \mbox{\Q@Bird@}.
Again, there are many reasons this would be bad, the most dramatic is that it would not scale to a scenario when the programmer of \Q@Bird@ and the programmer of \Q@Point3D@ are different.
\item Redefine getters and setters in \Q@Bird@, always put \Q@Point3D@ objects in the field and cast
the value out of the \Q@Point2D@ field to \Q@Point3D@ when implementing the overridden getter.
This solution scales to the multiple programmers approach, but requires ugly casts and can be implemented in a wrong way leading to bugs.
\end{itemize}

@@

\paragraph{Field type refinement with object interfaces}
Object interfaces address the challange of type-refinement as follows:
\begin{itemize}
\item by covariant method overriding we refine the type of the location field to \texttt{Point3D};
\item by \emph{overloading} we define a new setter for the location field with the more precise type;
\item using a default method we provide an implementation for the setter called 
with the old signature.
\end{itemize}

\lstinputlisting[linerange=68-72]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

@@

\paragraph{\Q@Point3D@ and properties updater}
To implement the old setter in a convenient way, \mixin supports one
last type of operations: property updater \texttt{with}
methods. Unlike the \texttt{withX} (where \texttt{X} stands for a
field name) methods presented so far, property updaters take several
fields at once, contained in an interface, and copy those fields into
fields of another interface.
The \Q@Point3D@ interface is defined as follows:
\lstinputlisting[linerange=51-56]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D
\noindent By using the new \Q@with@ method we may use the information for
\Q@z@ already stored in the object to forge an appropriate \Q@Point3D@
to store.  Note how all the informations about what fields sits in
\Q@Point3D@ and what in \Q@Point2D@ is properly encapsulated in the
\Q@with@ method, and is transparent for the implementer of \Q@Bird@.

\paragraph{Generated boilerplate}
Just to give a feeling on how much boring code \mixin is generating,
we show the generated code for the \texttt{Point3D} interface 
Figure~\ref{fig:boilerplate}. The generated code is very repetitive,
writing such code by hand can easily induce bugs, for example a
distracted programmer may swap the arguments of one of the many calls of
\Q@Point3D.of@.
\begin{figure}
\begin{lstlisting}
interface Point3D extends Point2D{
    Point3D withX(int val); Point3D withY(int val); Point3D withZ(int val);
    Point3D with(Point2D val);
    public static Point3D of(int _x, int _y, int _z){
        int x=_x;    int y=_y;    int z=_z;
        return new Point3D(){
            public int x(){return x;}    public int y(){return y;}    public int z(){return z;}
            public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
            public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
            public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
            public Point3D with(Point2D val){
                if(val instanceof Point3D){return (Point3D)val;}
                return Point3D.of(val.x(),val.y(),this.z()); }};}}
\end{lstlisting}
\caption{Generated boilerplate code.}
\label{fig:boilerplate}
\end{figure}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin does not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.

@@

\lstinputlisting[linerange=76-76]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=PEGASUS

@@

\lstinputlisting[linerange=51-56]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D

\lstinputlisting[linerange=59-59]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=ANIMAL

\lstinputlisting[linerange=63-64]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=HORSE

\lstinputlisting[linerange=68-72]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

\lstinputlisting[linerange=76-76]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=PEGASUS

>>>>

\begin{figure}
\begin{tabular}{c|c}
\includegraphics[height=4cm]{PegasusDetail.pdf}\hspace{20pt} &
\begin{minipage}{7cm}
\vspace{-90pt}
\lstinputlisting[linerange=6-17]{../UseMixinLombok/src/pegasusJava/Main.java}% APPLY:linerange=PEGASUS_JAVA
%basicstyle=\ttfamily\scriptsize
\end{minipage}
\end{tabular}
\caption{The animal system. Complete animal system on the left, and code for simplified animal system on the right. }\label{fig:pegasus}
\end{figure}

@@

\vspace{-5pt}
\paragraph{\Q@Point2D@ with Object Interfaces}
This obvious constructor code can be automatically generated by our \mixin
annotation.  By annotating the interface \Q@Point2D@, a variation of the shown
static method \texttt{of} will be generated, mimicking the functionality of a
simple-minded constructor: by looking at the abstract methods it first detects
what are the fields, then generates an \Q@of@ method with one parameter for each
of them. That is, we can just write

@@

\noindent More precisely, a field or factory parameter is generated for every
abstract method that takes no parameters (except for methods with special
names). An example of using \Q@Point2D@ is:
\begin{lstlisting}
Point2D p = Point2D.of(42,myPoint.y());
\end{lstlisting}
\noindent where we return a new point, using \Q@42@ as x-coordinate,
and taking all the other informations (only \Q@y@ in this case) from
another point.

\paragraph{\texttt{with} methods in Object Interfaces}
The pattern of creating a new object by reusing most information from an old
object is very common when programming with immutable data-structures. It is
supported in our code generation as \Q@with@ methods, for example:
\begin{lstlisting}
@Obj interface Point2D { 
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); // with methods }
\end{lstlisting}

\noindent Using \texttt{with} methods, the point \texttt{p} can be created instead using:

@@

\noindent If there is a large number of fields, \texttt{with} methods
will save programmers from writing large amount of tedious code that
simply copies field values. 
\begin{comment}
is expanded by our annotation into\footnote{
Note how we actually generate a real field \Q@int x=_x;@.
This provide a more uniform translation that can work also for mutable datastructures, where setters are required.
}

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
\end{comment}
Moreover, if the programmer wants a different implementation, they may 
just provide new \texttt{with} methods using \Q@default@:
\begin{lstlisting}
@Obj interface Point2D{ 
  int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}

@@

\paragraph{\Q@Animal@ and \Q@Horse@: simple mutable data with fields}
2D points are mathematical entities, thus we choose immutable data structure to
model them. However animals are real world entities, and when an animal moves,
it is the \emph{same} animal with a different location. We model this with
mutable state.

%Now we proceed to define \texttt{Animal} with \texttt{point} ``member
%field''.
%%Not again, before we used only getters!
% Again, we model this member field with getter and setter methods:
\lstinputlisting[linerange=57-57]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=ANIMAL

\noindent Here we declare abstract getter and setter for the mutable ``field''
\Q@location@.  Without the \mixin annotation, there is no convenient way to
instantiate \texttt{Animal}.  For \texttt{Horse}, the \mixin annotation is used
and an implementation of \texttt{run()} method is defined using \Q@default@,
which further illustrates the convenience of \emph{with} methods:

\lstinputlisting[linerange=61-62]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=HORSE

\noindent Creating and using \texttt{Horse} is quite simple:

@@

\paragraph{Summary} Dealing with states (mutable or not) in object interfaces
relies on a notion of abstract state, where only methods that interact with
states are available to users. Object interfaces provide support for four
different types of abstract state operations, which are summarized in
Figure~\ref{fig:abstractstate}.  The abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
mutable setters, and are discussed in detail in Section~\ref{sec:dsls}.

@@

\subsection{Object Interfaces and Subtyping}
\Q@Bird@s are \Q@Animal@s, but while \Q@Animal@s only need 2D
locations, \Q@Bird@s need 3D locations. Therefore when the \texttt{Bird}
interface extends the animal interface, the notion of points needs to
be refined. Such kind of refinement usually poses a challenge
in typical class-based approaches. Fortunatelly, with object interfaces,
we are able to provide a simple and effective solution.

\paragraph{Unsatisfactory class-based solutions to field type refinement}
In Java if we define an animal class with a field we have a set of
unsatisfactory options in front of us:

\begin{itemize}

\item Define a \Q@Point3D@ field in \Q@Animal@: this is bad since all animals
  would require more than needed.
  %Also it requires the programmer to predict the future, or 
  Also it requires adapting the old code to accommodate for new evolutions.

\item Define a \Q@Point2D@ field in \Q@Animal@ and define an extra \Q@int z@
  field in \Q@Bird@.  This solution is very ad-hoc, requires to basically
  duplicate the difference between \Q@Point2D@ and \Q@Point3D@ inside \Q@Bird@.
  %Again, there are many reasons this would be bad, 
  The most dramatic criticism is that it would not scale to a scenario when
  \Q@Bird@ and \Q@Point3D@ are from different programmers.

\item Redefine getters and setters in \Q@Bird@, always put \Q@Point3D@ objects
  in the field and cast the value out of the \Q@Point2D@ field to \Q@Point3D@
  when implementing the overridden getter.  This solution scales to the multiple
  programmers approach, but requires ugly casts and can be implemented in a
  wrong way leading to bugs.
\end{itemize}

@@

\paragraph{Field type refinement with object interfaces}
Object interfaces address the challenge of type-refinement as follows:
\begin{itemize}
\item by \emph{covariant method overriding} the return type of
  \texttt{location()} is refined to \texttt{Point3D};
\item by \emph{overloading} a new setter for location is defined with a more
  precise type;
\item we provide a \Q@default@ setter with the old signature.
\end{itemize}

\lstinputlisting[linerange=66-70]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

@@

\paragraph{\Q@Point3D@ and properties updater}
The \Q@Point3D@ interface is defined as follows:

\lstinputlisting[linerange=51-54]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D

\noindent By using the new \Q@with@ method we may use the information for
\Q@z@ already stored in the object to forge an appropriate \Q@Point3D@
to store.  Note how all the informations about what fields sits in
\Q@Point3D@ and what in \Q@Point2D@ is properly encapsulated in the
\Q@with@ method, and is transparent for the implementer of \Q@Bird@.

\Q@with-@ methods (functionally) update a field at a time.  This can be
inefficient, and sometime hard to maintain.  Often we want to update multiple
fields simultaneously, for example using another object as source.  Following
this idea, the method \Q@with(Point2D)@ is an example of a (functional)
properties updater: it takes a certain type of object and returns a copy of the
current object with all fields that match the fields of parameter type
updated. The idea is that we want the result to be still like \Q@this@, but
modified as similar as possible to the parameter.

% To implement the old setter in a convenient way, \mixin supports one
% last type of operations: property updater \texttt{with}
% methods. Unlike the \texttt{withX} (where \texttt{X} stands for a
% field name) methods presented so far, property updaters take several
% fields at once, contained in an interface, and copy those fields into
% fields of another interface.

\paragraph{Generated boilerplate}

\begin{figure}
\lstinputlisting[linerange=79-91]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=GENERATED_POINT3D
\caption{Generated boilerplate code.}
\label{fig:boilerplate}
\end{figure}

Just to give a feeling on how much boring code \mixin is generating, we show the
generated code for the \texttt{Point3D} in Figure~\ref{fig:boilerplate}. The
generated code is very mechanical, writing such code by hand is error-prune. For
example a distracted programmer may swap the arguments of calls to
\Q@Point3D.of@.  Note how \Q@with-@ methods are automatically refined in their
return type, so that code like \Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@
will be accepted and behave as expected. If the programmer wishes to suppress
this behaviour and keep the signature as it was, it is sufficient to redefine
the \Q@with-@ methods in the new interface repeating the old signature.  Again,
the philosophy is that if the programmer provides something directly, \mixin
does not touch it.  The cast in \Q@with(Point2D)@ is trivially safe since it is
guarded by an \Q@instanceof@ test. The idea is that if the parameter is a
subtype of the current exact type, then we can just return the parameter, as
something that is just ``more'' than \Q@this@.

@@

\lstinputlisting[linerange=74-74]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=PEGASUS

@@

\lstinputlisting[linerange=51-54]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D

\lstinputlisting[linerange=57-57]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=ANIMAL

\lstinputlisting[linerange=61-62]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=HORSE

\lstinputlisting[linerange=66-70]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

\lstinputlisting[linerange=74-74]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=PEGASUS
